Type Classes for Mathematics in Type Theory

### 2010-12-13 papers/mscs/mscs.tex
Adding references.

% \documentclass{mscs}
\documentclass[a4paper,10pt,runningheads]{llncs}
\usepackage{url}
\usepackage{bbm}
\usepackage{amsmath}
\usepackage{harvard}
\usepackage{amssymb}
\usepackage{upgreek}
\usepackage[mathletters]{ucs}
\usepackage[utf8x]{inputenx}
\usepackage{comment}
\newcommand{\N}{\ensuremath{\mathbbm{N}}}
\newcommand{\Z}{\ensuremath{\mathbbm{Z}}}
\newcommand{\Q}{\ensuremath{\mathbbm{Q}}}

@@

Our development is not merely aimed at formalization of theory, and our choice of a system based on type theory is no accident. It is our explicit ambition for the interfaces and theory we develop to be employed directly for the specification and parameterization of efficiently executable implementations of structures and procedures, expressed using type theory's native term reduction as a programming language. Thus, our work very much belongs in the long tradition of trying to realize to the greatest possible extent the promise of type theory to truly unite mathematics and certified (functional) programming, without making painful sacrifices on either side.

@@

 In section~\ref{classes}, we discuss our algebraic hierarchy implemented using type classes. In sections~\ref{cat} and~\ref{univ} we give a taste of what category theory and universal algebra look like in our development, and in section~\ref{numbers} we use these facilities to build abstract interfaces for numbers. In order to illustrate a very different use of type classes, we discuss the implementation of a quoting function for algebraic terms in terms of type classes, in section~\ref{quoting}. In section~\ref{sequences} we hint at an interface for sequences, but describe how a limitation in the current implementation of Coq makes its use problematic. We end with conclusions in section~\ref{conclusions}.

@@

Thus, bundled representations come at a substantial cost in flexibility. Historically, using bundled representations has nevertheless been an acceptable trade-off, because (1) the unbundled alternative was such a pain, and (2) the standard algebraic hierarchy (up to, say, fields and modules) is not all that wild.

@@

With these in place, our type class for categories follows the usual type-theoretical definition of a
category~\cite{saibi1995constructive}:
\begin{lstlisting}
  Class Category (O: Type) `{Arrows O} `{∀$$x y: O, Equiv (x ⟶ y)}
      `{CatId O} `{CatComp O}: Prop :=
    { arrow_equiv:> ∀ x y, Setoid (x ⟶ y)
    ; comp_proper:> ∀ x y z, Proper (equiv ==> equiv ==> equiv) comp
    ; comp_assoc w x y z (a: w ⟶ x) (b: x ⟶ y) (c: y ⟶ z):
        c ◎$$ (b ◎$$ a) = (c ◎$$ b) ◎$$ a
    ; id_l `(a: x ⟶ y): cat_id ◎$$ a = a
    ; id_r `(a: x ⟶ y): a ◎$$ cat_id = a }.
\end{lstlisting}
This definition is based on the 2-categorical idea of having equality only on arrows, not on objects.
%Similarly, we will have equality on natural transformations, but not on functors. % todo: ORLY?

@@

This work has been partially funded by the FORMATH project, nr.\ 243847, of the FET program within the 7th Framework program of the European Commission.
% MSCS recommends harvard, but does not specify which bst file to use.
\bibliographystyle{agsm}
\bibliography{alg}
\end{document}

>>>>

% \documentclass{mscs}
\documentclass[a4paper,10pt,runningheads]{llncs}
\usepackage{url}
\usepackage{bbm}
\usepackage{amsmath}
% \usepackage{harvard}
\usepackage{amssymb}
\usepackage{upgreek}
\usepackage[mathletters]{ucs}
\usepackage[utf8x]{inputenx}
\usepackage{comment}
\newcommand{\N}{\ensuremath{\mathbbm{N}}}
\newcommand{\Z}{\ensuremath{\mathbbm{Z}}}
\newcommand{\Q}{\ensuremath{\mathbbm{Q}}}

@@

Our development is not merely aimed at formalization of theory, and our choice of a system based on type theory is no accident. It is our explicit ambition for the interfaces and theory we develop to be employed directly for the specification and parameterization of efficiently executable procedures and data structures, implemented using type theory's native term reduction as a programming language. Thus, our work very much belongs in the long tradition of trying to realize the promise of type theory to truly unite mathematical formalization and certified (functional) programming, without making painful sacrifices on either side.

@@

 In section~\ref{classes}, we discuss our algebraic hierarchy implemented with type classes. In sections~\ref{cat} and~\ref{univ} we give a taste of what category theory and universal algebra look like in our development, and in section~\ref{numbers} we use these facilities to build abstract interfaces for numbers. In order to illustrate a very different use of type classes, we discuss the implementation of a quoting function for algebraic terms in terms of type classes, in section~\ref{quoting}. In section~\ref{sequences} we hint at an interface for sequences, but describe how a limitation in the current implementation of Coq makes its use problematic. We end with conclusions in section~\ref{conclusions}.

@@

Thus, bundled representations come at a substantial cost in flexibility. Historically, using bundled representations has nevertheless been an acceptable trade-off, because (1) the unbundled alternative was such a pain, and (2) the standard algebraic hierarchy (up to, say, fields and modules) is not all that wild. Further instances of the cost of sharing can be found in Section~\ref{classes}.

@@

With these in place, our type class for categories follows the usual type-theoretical definition of a
category~\cite{saibi1995constructive,PhdSAIBI}:
\begin{lstlisting}
  Class Category (O: Type) `{Arrows O} `{∀$$x y: O, Equiv (x ⟶ y)}
      `{CatId O} `{CatComp O}: Prop :=
    { arrow_equiv:> ∀ x y, Setoid (x ⟶ y)
    ; comp_proper:> ∀ x y z, Proper (equiv ==> equiv ==> equiv) comp
    ; comp_assoc w x y z (a: w ⟶ x) (b: x ⟶ y) (c: y ⟶ z):
        c ◎$$ (b ◎$$ a) = (c ◎$$ b) ◎$$ a
    ; id_l `(a: x ⟶ y): cat_id ◎$$ a = a
    ; id_r `(a: x ⟶ y): a ◎$$ cat_id = a }.
\end{lstlisting}
This definition is based on the 2-categorical idea of having equality only on arrows, not on objects.
%Similarly, we will have equality on natural transformations, but not on functors. % todo: ORLY?

@@

This work has been partially funded by the FORMATH project, nr.\ 243847, of the FET program within the 7th Framework program of the European Commission.
% MSCS recommends harvard, but does not specify which bst file to use.
\bibliographystyle{plain}
\bibliography{alg}
\end{document}
